Sale difference between Loblaws and Voila

### 2024-11-15 paper/paper.qmd
Update paper.qmd

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

```{r}
#| echo: false
#| warning: false
#| message: false
print(data)
```

```{r}
#| echo: false
#| warning: false
#| message: false
data_long <- data %>%
  mutate(price_type = ifelse(!is.na(other), "Sale", "Regular")) %>%
  pivot_longer(cols = c(current_price, old_price), values_to = "price")

# Create the line plot
ggplot(data_long, aes(x = product_vendor, y = price, color = price_type, linetype = price_type, group = interaction(price_type, product_vendor))) +
  geom_line() +
  geom_point() +
  facet_wrap(~product_vendor) + # Separate lines for each vendor
  labs(title = "Price Comparison Over Time for Apples Ambrosia",
       x = "Vendor", 
       y = "Price", 
       color = "Price Type", 
       linetype = "Price Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Measurement
To investigate the potential for artificial price increases prior to sales events, this study employs a data-driven measurement approach to track and analyze price fluctuations over time. The measurement strategy is broken down as follows:

@@

# Results
Price Analysis of "Apples Ambrosia" Across Vendors
The dataset consists of price data for Apples Ambrosia from two grocery vendors: Voila and Loblaws. For each vendor, prices are observed during periods labeled as "SALE" (Voila) and regular pricing periods (Loblaws). The analysis investigates the following:

Price Comparison Between Vendors:

Voila offers a current price for the apples ranging from $5.49 to $5.99, with an old price of $6.49 for all sale items. This suggests a $1.00 decrease in price during the sale period, indicating that Voila is promoting a sale by reducing the price from the regular price.
Loblaws, on the other hand, has a consistent current price of $5.49 for all data points. This is lower than the sale price at Voila. Additionally, Loblaws’ old price is slightly higher, averaging around $5.49 to $6.58. However, the regular price at Loblaws appears to be stable without the promotional fluctuations observed at Voila.
Sale Behavior Analysis at Voila:

At Voila, the old price for the apples is $6.49, which is advertised as the regular price. The current price during the sale is lower by exactly $1.00, consistently across multiple observations. This consistent pricing pattern across the dataset suggests that Voila uses a fixed price reduction strategy during sales, offering a price decrease of $1.00 off the original price during promotions.
An important observation is that the old price at Voila (before the sale) is consistently $6.49, which may suggest that prices were artificially elevated just ahead of the sale, only to be reduced back down to the normal sale price during the promotional period. The stable difference between the sale and non-sale prices raises a question about the pricing practices at Voila.
Price Consistency at Loblaws:

At Loblaws, no specific sale event is recorded, and the prices are relatively stable, with current prices consistently at $5.49 for "Apples Ambrosia." The old price for Loblaws varies slightly, but the regular price does not undergo significant fluctuations like the sales at Voila. This stability in pricing indicates that Loblaws may not engage in price manipulation strategies for sale periods as seen at Voila. The current price of $5.49 is relatively consistent and does not appear to be influenced by any temporary sale-like conditions.
Conclusion on Price Manipulation:

Based on the price data, Voila may be engaging in price manipulation practices, where the price of the apples is first elevated to $6.49 and then temporarily reduced to $5.49 during the sale period, creating an illusion of a larger discount. This could be viewed as an artificial price increase before a sale, which aligns with the research question regarding whether prices are jacked up before a sale and lowered back to normal.
Loblaws, by contrast, maintains a more stable pricing structure and does not appear to engage in such pricing practices.
Overall, this analysis suggests that Voila may be artificially inflating prices ahead of sales to make the sale price appear more significant, while Loblaws maintains a consistent price strategy without such fluctuations. Further statistical analysis, such as testing for significant price changes before and after sales, would be necessary to substantiate this hypothesis more robustly.

@@

## Conclusion

In conclusion, the analysis indicates that **Voila** may be engaging in artificial price elevation ahead of sales, creating an illusion of a larger discount when the price is lowered back to normal during the sale period. **Loblaws**, in contrast, maintains a consistent and stable pricing strategy. While this raises important questions about transparency and fairness in pricing practices, it also provides an opportunity for consumers to make more informed decisions by understanding how sale prices may be manipulated. Further research would be needed to investigate this practice across a broader range of products and vendors, with an emphasis on consumer perceptions and market dynamics.

>>>>

## Overview
We use the statistical programming language R [@citeR]. The data used in this paper came from the [@hammer]. Following @tellingstories, we considerto capture the pricing dynamics of Ambrosia apples across different vendors, we extracted relevant data from a larger dataset containing multiple products and vendors. The initial dataset included information from seven vendors, each offering various products. Our focus was on the specific product, "Ambrosia Apples," which was sold by two vendors—NoFrills and Loblaws.From this comprehensive dataset, we filtered out all other products, isolating only the entries related to Ambrosia apples. The extracted data included key information such as the product ID, product name, current price, and old price. Additionally, a field labeled "other" contained further notes on sale conditions or discounts. The dataset also recorded timestamps, allowing us to track pricing changes over time.

## Measurement

To investigate the potential for artificial price increases prior to sales events, this study employs a data-driven measurement approach to track and analyze price fluctuations over time. The measurement strategy is broken down as follows:

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-loblaws_data
#| tbl-cap: "Ambrosia Apples Data of Loblaws"
#| warning: false

data |>
  filter(product_vendor == "Loblaws") |>
  select(product_name,product_vendor,current_price,old_price,nowtime) |>      # Select the first 3 columns
  kable(
    col.names = c("Product Name","Product Vendor","Product Price(current)","Product Price(old)","Date"),
    booktabs = TRUE
  )
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-nofrills_data
#| tbl-cap: "Ambrosia Apples Data of NoFrills"
#| tbl-width: 2
#| tbl-height: 2
#| warning: false

data |>
  filter(product_vendor == "NoFrills") |>
  select(product_name,product_vendor,current_price,old_price,nowtime) |>    
  kable(
    col.names = c("Product Name","Product Vendor","Product Price (current)","Product Price (old)","Date"),
    booktabs = TRUE
  )
```

The transition from real-world phenomena (e.g., pricing decisions and promotional strategies) to dataset entries was facilitated by these fields.Current and old prices reflected the vendors' listed prices at the time of data collection. Notes in the "other" field provided contextual information about whether the price was marked as a sale or if there were discounts. The "nowtime" column captured the exact timing of data collection, enabling us to analyze price fluctuations and assess sale validity over time.

By focusing on two vendors and this specific product, we created a clean, targeted dataset shows in @tbl-loblaws_data and @tbl-nofrills_data that accurately reflects the pricing behavior for Ambrosia apples at NoFrills and Loblaws, laying the foundation for further analysis.

## Outcome variables

```{r}
#| label: fig-compare_price_vendor
#| fig-cap: Current and Old Prices of Ambrosia Apples Over Time
#| echo: false
#| warning: false
#| message: false

clean_data$nowtime <- as.POSIXct(clean_data$nowtime, format="%Y-%m-%d %H:%M:%S")

clean_data$product_vendor <- factor(clean_data$product_vendor, levels = c("NoFrills", "Loblaws"))

# Create the line plot
ggplot(clean_data, aes(x = nowtime)) +
  geom_line(aes(y = current_price, color = product_vendor, linetype = "Current Price")) +
  geom_line(aes(y = old_price, color = product_vendor, linetype = "Old Price")) +
  labs(title = "Current and Old Prices of Ambrosia Apples Over Time",
       x = "Date",
       y = "Price ($)",
       color = "Vendor",
       linetype = "Price Type") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme(legend.position = "right")
```
We compared the apple prices of nofrills and loblaws at different times (including the comparison of old prices, where NoFrills does not include any old prices because this vendor does not have any discounts for this product and always maintains the original price), which can be seen in @fig-compare_price_vendor.

# Results
Price Analysis of "Apples Ambrosia" Across Vendors
The dataset consists of price data for Apples Ambrosia from two grocery vendors: NoFrills and Loblaws. For each vendor, prices are observed during periods labeled as "SALE" (Loblaws) and regular pricing periods (NoFrills). The analysis investigates the following:

## Price Comparison Between Vendors
Loblaws offers a current price for the apples \$1.21, with an old price of \$1.45 for all sale items. This suggests a $0.24 decrease in price during the sale period, indicating that Loblaws is promoting a sale by reducing the price from the regular price.
NoFrills, on the other hand, has a consistent current price of $1.2 for all data points. This is the same as the sale price at Loblaws. Additionally, NoFrills has no old price since they didn't do any discounts. However, the regular price at NoFrills appears to be stable without the promotional fluctuations observed at Loblaws.

## Conclusion: Price Manipulation Practices?
Despite Loblaws advertising a price drop from \$1.45 to \$1.21 as part of a "sale," the price at NoFrills is consistently set at \$1.21, without any promotional claims. This raises suspicions that Loblaws may be engaging in price manipulation practices, where the advertised sale price is simply a temporary reduction to match NoFrills' regular price, rather than offering a genuine discount.
These findings highlight potential concerns about transparency in pricing and suggest that consumers may be misled into believing they are receiving a larger discount than is actually being offered. Loblaws' "sale" price may not reflect a true reduction in value, but rather a strategic pricing move to attract customers by promoting the illusion of a better deal. This could be seen as a form of pricing manipulation, where the apparent discount is used to create the perception of a more significant bargain.
Consumers should be cautious when interpreting sales promotions and be aware of the actual price comparisons across different vendors to make informed purchasing decisions.

@@

## Correlation vs. Causation
In this analysis, we have observed a correlation between the sale price at Loblaws and the regular price at NoFrills, suggesting that both prices are set at $1.21 during the sale period. However, it is important to note that correlation does not imply causation. While the two prices may appear to be related, this analysis does not establish a causal relationship between Loblaws' price reduction and NoFrills' pricing strategy. For example, it is possible that Loblaws adjusted its price temporarily to match NoFrills' pricing rather than making a true price reduction. A more comprehensive causal analysis would require additional data points, such as customer behavior during sale periods, or experiments where prices are manipulated independently, to determine whether Loblaws' pricing strategy is indeed a response to NoFrills or if there are other factors influencing both vendors' pricing.

## Missing Data
Our dataset consists of prices from two vendors (NoFrills and Loblaws), but the full dataset includes many more products and vendors. Missing data could be a potential concern in our analysis, especially if key information, such as prices for other vendors, was not available or if data points for certain time periods were omitted. Missing data can bias the analysis if it is not random; for example, if sales data is missing during peak shopping periods when prices are likely to be highest, this could result in a lower estimate of the actual price difference. In our case, the absence of historical prices for NoFrills or detailed pricing data for other vendors might limit the conclusions we can draw about the broader pricing landscape. To address this, we could perform sensitivity analyses to evaluate how the results change with different assumptions about the missing data, or use imputation techniques to fill in gaps where appropriate.

## Sources of Bias
There are several potential sources of bias in this study that could influence our results. First, selection bias could occur if the data points from Loblaws and NoFrills are not representative of all pricing periods, or if certain time frames (e.g., weekends or holidays) are overrepresented. Since our data focuses on a subset of vendors and products, it may not fully capture the diversity of pricing practices across all 7 vendors in the full dataset. Additionally, vendor-specific biases may also exist if either Loblaws or NoFrills have particular promotional strategies that differ from those of other vendors. For instance, Loblaws' advertised sales may be selectively applied to high-demand products like Ambrosia apples, while NoFrills might use a more consistent pricing approach across its entire range of produce. Furthermore, potential reporting biases could affect the accuracy of the data if certain pricing changes were not accurately recorded or if vendor promotions were not consistently reflected in the dataset. To mitigate these biases, future analyses could involve a broader set of vendors and products, as well as methods to ensure that the dataset is representative and balanced.
